Comparison of Fast, Hybrid Imaging Architectures for Multi-scale, Hierarchical Aperture Arrays

### 2024-09-29 response_to_review_1.txt

--> I thank the reviewer for a thorough review of the manuscript. I have addressed the concerns as detailed below. My responses are preceded with "-->". Changes in the manuscript have been highlighted with bold font.

Comments to the Author
This paper addresses the important question of how to choose a computational architecture for a given aperture-array radio telescope depending on how its collecting area is deployed, and specifically in the context of some current/upcoming arrays that use hierarchical distribution of antennas. The analysis is new and useful to the community. I found much of the paper to be quite dense in its presentation which unfortunately detracts from its readability. I address one aspect of this in the major comments below.

Major comments
=============
- The paper gets dense (in terms of readability) at the start of Section 5 and I would suggest more care to ensure that the key equations are well described. For example, equation (1) is presented with essentially no context, and the notation is not well explained. Many of the indices are undefined (what is p? what is alpha? what is the asterisk?). This really holds throughout all of Sections 5 and 6. Even though it will lengthen the paper somewhat, I think it’s crucial that you review these approaches and their associated equations carefully, ensuring that they are framed usefully for the reader and that the notation is clarified.
- I found myself wondering if there are any links to real-world performance metrics that you can make. In Section 5.2 you mentioned that LWA has implemented EPIC. Certainly other techniques have been used before. Are there any measured performance figures that can be used to validate the model used in this paper?

Minor comments
==============

General:
- The name of SKA and SKA-Low should be written differently throughout. (a) the "SKA" acronym is not meant to be expanded anymore - it's just SKA, and (b) "SKA" should be used in place of "SKA1" - the SKA1 vs SKA2 distinction isn't used anymore either. See for example: https://www.skao.int/en/skao-brand

Per section:

Title
- There's a misspelled word! (Hybrid)

Abstract
- I think the readability of the abstract could be improved. For example, the term "large-N" is a bit of jargon and I think the phrasing around this could be changed to more natural language.
- The use of "spatial scale" is similarly a bit jarring when first reading the paper. I think it's because the wording would more clearly be, for example, "...multiple spatial scales spanning the dimensions of individual elements, the size of stations (...), and the spacing between stations." (or similar)

Section 1
- First sentence, I think "polarizations" should be changed to "polarization states"
- Later in the first paragraph, when mentioning the various classes of transient/variable sources, I think it would be good to include a representative range of timescales (duration and/or periodicity and/or duty cycle). Perhaps not very lengthy, but to give a sense of relevant timescales to be considered later in the paper. (This is included much later in Section 7.3 but should I think be included at some level here in the Introduction too.)
- Similarly at the end of the second paragraph I think you should give some indicative values for "large-scale structures" and "high sensitivity".

Section 2
- I'm a bit confused by the choice of notation here. If you're attempting to be more general than "antenna", hence using the phrase "collecting element", why not use N_e for number of them? And the N_{a/s_m} construction seems quite awkward, is there a convenient alternative?
- At the end of that paragraph, it's good to note that usually all stations have the same number of antennas, but maybe you could note that e.g. LOFAR would not be so easily simplified.
- Is there a webpage or paper to refer to for LAMBDA-I ?
- Real-time imaging is assumed (last paragraph of this section) but I'm not aware that this is required for cosmological applications, e.g. EoR use cases typically make use of post-observation analysis and imaging. Does this undermine any of the arguments? Alternatively, if I've missed something, could information be included to justify why this is needed?

Section 3
- Detailed point, should "per pixel" (first paragraph) really be "per resolution element" or similar?
- In the third paragraph, it seems like a major 'other cost' is calibration. Later in the paper you make the argument that calibration is low-cost and also carried out on significantly slower time scales (at least relevant to fast imaging for transients). Possibly add this sort of consideration here as well. And possibly note here too that it's assumed in Section 5 that calibration has already occurred (for example this seems clearly an assumption at the start of Section 5.1).

Section 4
- First paragraph, the sentence starting "The second stage..." doesn't make sense, can this be reworded?

Figure 1
- "Accumulation" is misspelled in 6(?) places

Section 5
- In the first paragraph after equation (3), I think there’s an asterisk missing in the inline equation embedded in the sentence starting “For example, choosing…”
- In the next paragraph, I think “can hold any N_{a/s}” should be “can hold as many N_{a/s}”
- In the next sentence, can you provide more about the constant padding factor (what is its role)?

Section 6
- In 6.2.4, it struck me that there is no consideration about the impact of the w-term. I think the highly formalised way that equation (13) is presented may be elegant but it obfuscates the practical aspects that radio astronomers are familiar with, like dealing with the w-term. More generally there are many ways that this procedure can be implemented in practice. For example the gridding can be managed various ways (including image plane gridding). What is the assumption used here as described in the context of those practical implementations?

Figures 2-5
- These refer to LAMBDA-I stations, but from Figure 6 onward you rightly start referring to SKA-Low as well since the stations for these designs are nominally the same. I suggest to make this consistent throughout the paper.

Figure 5a
- I’m not getting much from this diagram since the colorscale limits don’t allow whatever variation there is in the plotted values to come through.

Figures 2,4,6,8,10,11
- I’m not sure what is driving the range of the parameters that you’re plotting here, but it’s confusing and may hide information to have the axis limits significantly larger than the range of values for which there are data to plot. Why not either extend the range of values computed and plotted (if/where it makes sense) and/or reduce the axis limits to avoid orders of magnitude of blank space?

Figure 9
- The caption is incomplete (end of first sentence)

### 2024-10-02 ms.tex
Start addressing minor comments, update figure 1

\title{Comparison of Fast, Hybird Imaging Architectures for Multi-scale, Hierarchical Aperture Arrays}

@@

\begin{abstract}
Two major areas of modern radio astronomy, namely, explosive astrophysical transient phenomena and observations of cosmological structures, are driving aperture arrays towards large-$N$, low-cost element design consisting of multiple spatial scales comprising the individual elements, stations (a collection of elements), and a grouping of stations. Such multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different spatial scales to obtain optimal performance in imaging the entire field of view. Adopting a computational cost metric based on the number of floating point operations, its distribution over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time is examined to determine the most efficient hybrid architectures over the parameter space of hierarchical aperture array layouts. Nominal parameters of specific upcoming and planned arrays-- Square Kilometre Array (SKA) at low frequencies (\texttt{SKA1-low}), \texttt{SKA1-low-core}, a long baseline extension to \texttt{SKA1-low} (\texttt{LAMBDA-I}), compact all-sky phased array (\texttt{CASPA}), and a lunar array (\texttt{FarView-core})-- are used to determine the most optimal architecture hierarchy for each from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large-$N$, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval, which in turn is determined by the science goals.
\end{abstract}

@@

The variable and transient radio sky encompasses a variety of intriguing astrophysical phenomena with a range of timescales spanning orders of magnitude, wide frequency ranges, and polarizations \citep[][and references therein]{Pietka+2015,Chandra+2016,Nimmo+2022}. Notable examples include pulsars and their giant radio pulses, magnetars, fast radio bursts (FRBs), rotating radio transients (RRATs), Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, gamma-ray bursts (GRBs) and their afterglows, active galactic nuclei (AGNs) and blazars, tidal disruption events, both the prompt and afterglow counterparts to gravitational wave (GW) merger events, as well as many phenomena yet to be understood \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}. The effectiveness of discovering these transients depends on the instrument's field of view and the duration of on-sky observation \citep{Cordes2007}. Ideally, instruments should operate continuously and in real time, covering a wide range of timescales from microseconds to days, with a large instantaneous field of view.

@@

Based on the requirements for transients and observational cosmology, radio interferometers are witnessing a paradigm shift towards packing a large number of relatively small-sized collecting elements densely in a compact region. Modern radio arrays tend to be aperture arrays that operate interferometrically and hierarchically on multiple spatial scales, involving spatial dimensions of the collecting element, a spatial grouping of elements forming a virtual telescope typically referred to as a station or a tile, and a spatial grouping of stations comprising the full array.
Examples of multi-scale, hierarchical aperture arrays include the Murchison Widefield Array \cite[MWA;][]{Tingay+2013}, the Hydrogen Epoch of Reionization Array \citep[HERA;][]{HERA+2017}, the Low Frequency Array \cite[LOFAR;][]{vanHaarlem+2013}, the swarm of Long Wavelength Arrays \cite[LWA Swarm;][]{Dowell+2018}, 
% the Canadian Hydrogen Intensity Mapping Experiment \citep[CHIME;][]{CHIME+2022}, 
the Hydrogen Intensity and Real-time Analysis Experiment \citep[HIRAX;][]{HIRAX+2022}, and the Square Kilometre Array low \cite[SKA-low;][]{Dewdney+2009,SKA1+2019}.

@@

The specific aperture array examples considered are the \texttt{SKA1-low}, the core of the \texttt{SKA1-low} (\texttt{SKA1-low-core}), the first phase of a long baseline extension to \texttt{SKA1-low} called Low-frequency Australian Megametre Baseline Demonstrator Array (\texttt{LAMBDA-I}), the Compact All-Sky Phased Array \citep[\texttt{CASPA}\footnote{CASPA's stations are intnded for localisation and astrometry, rather than for aperture synthesis.};][]{Luo+2024}, and the core of a lunar array \citep[\texttt{FarView};][]{Polidan+2024} which is hereafter referred to as \texttt{FarView-core}. \texttt{LAMBDA-I}, \texttt{CASPA}, and \texttt{FarView-core} are hypothetical interferometers that are being considered for the future.

@@

\begin{table*}[htb!]
\normalsize
\begin{threeparttable}
\caption{Table of multi-scale interferometer array parameters.}
\label{tab:array_params}
\begin{tabular}{ccccccc|cc}
\toprule
\headrow Array name & $\lambda$ (m)\tnote{a} & $D_\textrm{A}/\lambda$ & $N_\textrm{s}$ & $D_\textrm{s}/\lambda$ & $N_\textrm{a/s}$ & $D_\textrm{a}/\lambda$ & $f_\textrm{A}$\tnote{b} & $f_\textrm{s}$\tnote{c} \\
\midrule
\texttt{LAMBDA-I} & 2 & $3.9\times 10^6$ & 4 & 17.5 & 256 & 1 &  $1.61\times 10^{-10}$ & 0.836 \\ 
\texttt{SKA1-low-core} & 2 & $5.25\times 10^{2}$ & 256 & 17.5 & 256 & 1 & 0.284 & 0.836 \\ 
\texttt{SKA1-low} & 2 & $3.5\times 10^4$ & 512 & 17.5 & 256 & 1 &  $1.28\times 10^{-4}$ & 0.836 \\ 
\texttt{CASPA} & 0.25 & $4.6\times 10^4$ & 3 & 8.08 & 65 & 0.5 &  $9.2\times 10^{-8}$ & 0.996 \\ 
\texttt{FarView-core}\tnote{d} & 10 & $6.77\times 10^2$ & 81 & 38.5 & 625 & 1 & 0.262 & 0.422 \\ 
\bottomrule
\end{tabular}
\begin{tablenotes}[hang]
% \item[]Table note
\item[a]Nominal operating wavelength of the array. Results may differ at other wavelengths depending on aperture efficiency, filling factor, etc. 
\item[b]Array filling factor, $\,f_\textrm{A}=N_\textrm{s}\,(D_\textrm{s}/D_\textrm{A})^2$
\item[c]Station filling factor, $\,f_\textrm{s}=N_\textrm{a/s}\,(D_\textrm{a}/D_\textrm{s})^2$
\item[d]Only 6~km core of the proposed array \citep{Polidan+2024} is considered here.
\end{tablenotes}
\end{threeparttable}
\end{table*}

@@

Two broad classes of two-stage fast imaging architectures are considered. In both classes, the first stage is intra-station interferometry of the element electric fields using the options of voltage beamforming (BF), E-field Parallel Imaging Correlator (EPIC), beamforming of correlations (XBF), or correlation followed by FFT (XFFT). The second stage inter-station combination of the data products from the first stage. The main difference between the two classes is in whether the second stage uses the first stage products coherently or incoherently. A schematic of these hybrid architectures and their computational cost components are illustrated in Figure~\ref{fig:hybrid-architectures-cost-schematic}.

@@

% \begin{figure*}
% \centering
% \subfloat[][Two-dimensional slices \label{fig:multidim-incoherent-compcost-BF-SKA1-low}]{\includegraphics[width=0.4\textwidth]{figures/multidim_beamformer_dft_cost_analysis_SKA1-low.pdf}}
% \subfloat[][One-dimensional slices\label{fig:1D-incoherent-compcost-BF-SKA1-low}]{\includegraphics[width=0.58\textwidth]{figures/1D_beamformer_dft_cost_analysis_SKA1-low.pdf}}
% \caption{(Left): Two-dimensional slices of the computational cost for coherent-incoherent imaging using station-level voltage beamforming for \texttt{SKA1-low-core}, \texttt{SKA1-low}, \texttt{Aus-VLBA-low}, and \texttt{IC-VLBA-low}. Cyan contours  \label{fig:incoherent-compcost-BF-SKA-low}}
% \end{figure*}

@@

Figure~\ref{fig:1D-incoherent-compcost-EPIC-LAMBDA-I} shows a breakdown of the costs per station per voxel of the EPIC architecture, with nominal values of a \texttt{LAMBDA-I} station marked in cyan. The dominant cost is from the two-dimensional FFT (black dashed lines) in Equation~\ref{eqn:intra-station-pol-hol-img-epic}.
% or equivalently, Equation~\ref{eqn:intra-station-pol-hol-img-epic-LA}.
Because the spatial FFT cost depends only on the grid dimensions, it is independent of $N_\textrm{a/s}$ as the grid can hold any $N_\textrm{a/s}$ elements as physically possible. This spatial FFT scales roughly as $(\gamma_\textrm{s} \, D_\textrm{s}/D_\textrm{a})^2\log_2(\gamma_\textrm{s}\, D_\textrm{s}/D_\textrm{a})^2$, where, $\gamma_\textrm{s}$ is a constant padding factor. Notably, because the spatial FFT produces an image over the entire field of view at every independent pixel instantly, it is not a pixelwise operation, and therefore, its cost per pixel is $\sim \log_2(D_\textrm{s}/D_\textrm{a})$, showing a weak dependence on $D_\textrm{s}$ and $D_\textrm{a}$. The pixelwise squaring (pink dot-dashed lines) and temporal averaging (blue dot-dashed lines) costs in Equation~\ref{eqn:intra-station-opt-pol-img-outprod}
% {eqn:intra-station-opt-polarimetric-direct-img-LA}
are subdominant. These being per pixel operations, the cost per voxel remains constant with $N_\textrm{a/s}$, $D_\textrm{s}$ and $D_\textrm{a}$ similar to the voltage beamforming architecture. The gridding (black dotted lines) with weights in Equation~(\ref{eqn:intra-station-pol-hol-img-epic})
% and (\ref{eqn:intra-station-pol-hol-img-epic-LA})
is a per-element operation and depends only on $N_\textrm{a/s}$, it scales linearly with $N_\textrm{a/s}$ and inverse squared with $D_\textrm{s}$ per voxel. Overall, the gridding cost is insignificant. Again, the spatial averaging of pixels across multiple stations (orange dot-dashed line) is not a station-level operation, but is only shown to emphasise that it operates on a much slower cadence of $t_\textrm{acc}$, scaling inversely with $t_\textrm{acc}$, and adds only negligible cost to the overall cost budget.

@@

\begin{figure}
\includegraphics[width=\linewidth]
% {figures/1D_xcor_dft_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_06.pdf}
\caption{A breakdown of the computational cost density function over the station parameters for the correlator beamforming architecture (XBF) at the station level. Each panel shows the variation with the respective parameter keeping the rest fixed at the characteristic values of the \texttt{LAMBDA-I}, \texttt{SKA1-low-core}, and \texttt{SKA1-low} stations (cyan lines). The two-dimensional DFT (dashed lines) dominates the cost for the chosen \texttt{LAMBDA-I} parameters at an imaging cadence of $t_\textrm{acc}=1$~ms over correlation (double dot-dashed) and temporal averaging of the correlations (dot-dashed).
% The spatial FFT cost in Equation~(\ref{eqn:intra-station-pol-hol-img-epic-LA}) denoted by black dashed lines dominates the squaring (pink dot-dashes) and temporal averaging (blue dot-dashes) costs in Equation~(\ref{eqn:intra-station-opt-polarimetric-direct-img-LA}). The gridding operation (black dotted lines) in Equation~(\ref{eqn:intra-station-pol-hol-img-epic-LA}) contributes even lesser to the cost budget. 
\label{fig:1D-incoherent-compcost-XBF-LAMBDA-I}}
\end{figure}

@@

\begin{figure*}
\centering
\subfloat[][Covariant cost density of XBF architecture \label{fig:multidim-incoherent-compcost-XBF-LAMBDA-I}]
{\includegraphics[width=0.48\textwidth]
% {figures/multidim_xcor_dft_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_07a.pdf}
}
\subfloat[][Covariant cost of XBF architecture relative to BF architecture \label{fig:multidim-incoherent-relative-compcost-XBF-LAMBDA-I}]
{\includegraphics[width=0.48\textwidth]
% {figures/multidim_xcor_dft_relative_cost_analysis_LAMBDA-I.pdf}
{fig_07b.pdf}
}
\caption{(Left): Two-dimensional slices of the computational cost for coherent-incoherent imaging using a DFT of station-level cross-correlations for \texttt{LAMBDA-I}, \texttt{SKA1-low-core}, and \texttt{SKA1-low} station parameters (dashed gray lines). Cyan lines denote contours of the colour scale in logarithmic increments. (Right): Same as the left but relative to a voltage beamformer architecture. The computational cost density is dominated by the two-dimensional DFT cost for $t_\textrm{acc}=1$~ms, and is lower than that of voltage beamforming for $N_\textrm{a/s}\lesssim 100$ while getting more expensive with increasing $N_\textrm{a/s}$. \label{fig:incoherent-compcost-XBF-LAMBDA-I}}
\end{figure*}

@@

% \begin{figure*}
% \centering
% \subfloat[][Two-dimensional slices \label{fig:multidim-incoherent-compcost-XBF-SKA1-low}]{\includegraphics[width=0.48\textwidth]{figures/multidim_xcor_dft_cost_analysis_SKA1-low.pdf}}
% \subfloat[][Two-dimensional slices (relative)\label{fig:multidim-incoherent-relative-compcost-XBF-SKA1-low}]{\includegraphics[width=0.48\textwidth]{figures/multidim_xcor_dft_relative_cost_analysis_SKA1-low.pdf}}
% \caption{(Left): Two-dimensional slices of the computational cost for coherent-incoherent imaging using a DFT of station-level cross-correlations for \texttt{SKA1-low-core}, \texttt{SKA1-low}, \texttt{Aus-VLBA-low}, and \texttt{IC-VLBA-low}. Cyan contours  \label{fig:incoherent-compcost-XBF-SKA1-low}}
% \end{figure*}

% \begin{figure}
% \includegraphics[width=\linewidth]{figures/1D_xcor_dft_cost_analysis_SKA1-low.pdf}
% % \includegraphics[width=\linewidth]{fig_01.pdf}
% \caption{
% \label{fig:1D-incoherent-compcost-XBF-SKA1-low}}
% \end{figure}

@@

\begin{figure}
\includegraphics[width=\linewidth]
% {figures/1D_xcor_fft_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_08.pdf}
\caption{A breakdown of the computational cost density function over the station parameters for the correlator FFT architecture (XFFT) at the station level. Each panel shows the variation with the respective parameter keeping the rest fixed at the characteristic values of the \texttt{LAMBDA-I}, \texttt{SKA1-low-core}, and \texttt{SKA1-low} stations (cyan lines). The correlator / X-engine (double dot- dashed lines) and gridding (dotted lines) dominate the cost for the chosen \texttt{LAMBDA-I} parameters at an imaging cadence of $t_\textrm{acc}=1$~ms over the temporal averaging (dot dahsed lines) and two-dimensional FFT (dashed lines).
\label{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}}
\end{figure}

>>>>

\title{Comparison of Fast, Hybrid Imaging Architectures for Multi-scale, Hierarchical Aperture Arrays}

@@

\begin{abstract}
Two major areas of modern radio astronomy, namely, explosive astrophysical transient phenomena and observations of cosmological structures, are driving aperture arrays towards large-$N$, low-cost element design consisting of multiple spatial scales comprising the individual elements, stations (a collection of elements), and a grouping of stations. Such multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different spatial scales to obtain optimal performance in imaging the entire field of view. Adopting a computational cost metric based on the number of floating point operations, its distribution over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time is examined to determine the most efficient hybrid architectures over the parameter space of hierarchical aperture array layouts. Nominal parameters of specific upcoming and planned arrays-- the SKA at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to \texttt{SKA-low} (\texttt{LAMBDA-I}), compact all-sky phased array (\texttt{CASPA}), and a lunar array (\texttt{FarView-core})-- are used to determine the most optimal architecture hierarchy for each from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large-$N$, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval, which in turn is determined by the science goals.
\end{abstract}

@@

The variable and transient radio sky encompasses a variety of intriguing astrophysical phenomena with a range of timescales spanning orders of magnitude, wide frequency ranges, and \textbf{polarisation states} \citep[][and references therein]{Pietka+2015,Chandra+2016,Nimmo+2022}. Notable examples include pulsars and their giant radio pulses, magnetars, fast radio bursts (FRBs), rotating radio transients (RRATs), Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, gamma-ray bursts (GRBs) and their afterglows, active galactic nuclei (AGNs) and blazars, tidal disruption events, both the prompt and afterglow counterparts to gravitational wave (GW) merger events, as well as many phenomena yet to be understood \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}. The effectiveness of discovering these transients depends on the instrument's field of view and the duration of on-sky observation \citep{Cordes2007}. Ideally, instruments should operate continuously and in real time, covering a wide range of timescales from microseconds to days, with a large instantaneous field of view.

@@

Based on the requirements for transients and observational cosmology, radio interferometers are witnessing a paradigm shift towards packing a large number of relatively small-sized collecting elements densely in a compact region. Modern radio arrays tend to be aperture arrays that operate interferometrically and hierarchically on multiple spatial scales, involving spatial dimensions of the collecting element, a spatial grouping of elements forming a virtual telescope typically referred to as a station or a tile, and a spatial grouping of stations comprising the full array.
Examples of multi-scale, hierarchical aperture arrays include the Murchison Widefield Array \cite[MWA;][]{Tingay+2013}, the Hydrogen Epoch of Reionization Array \citep[HERA;][]{HERA+2017}, the Low Frequency Array \cite[LOFAR;][]{vanHaarlem+2013}, the swarm of Long Wavelength Arrays \cite[LWA Swarm;][]{Dowell+2018}, 
% the Canadian Hydrogen Intensity Mapping Experiment \citep[CHIME;][]{CHIME+2022}, 
the Hydrogen Intensity and Real-time Analysis Experiment \citep[HIRAX;][]{HIRAX+2022}, and the \texttt{SKA-low} \cite{Dewdney+2009,SKA1+2019}.

@@

The specific aperture array examples considered are the \texttt{SKA-low}, the core of the \texttt{SKA-low} (\texttt{SKA-low-core}), the first phase of a long baseline extension to \texttt{SKA-low} called Low-frequency Australian Megametre Baseline Demonstrator Array (\texttt{LAMBDA-I}), the Compact All-Sky Phased Array \citep[\texttt{CASPA}\footnote{CASPA's stations are intnded for localisation and astrometry, rather than for aperture synthesis.};][]{Luo+2024}, and the core of a lunar array \citep[\texttt{FarView};][]{Polidan+2024} which is hereafter referred to as \texttt{FarView-core}. \texttt{LAMBDA-I}, \texttt{CASPA}, and \texttt{FarView-core} are hypothetical interferometers that are being considered for the future.

@@

\begin{table*}[htb!]
\normalsize
\begin{threeparttable}
\caption{Table of multi-scale interferometer array parameters.}
\label{tab:array_params}
\begin{tabular}{ccccccc|cc}
\toprule
\headrow Array name & $\lambda$ (m)\tnote{a} & $D_\textrm{A}/\lambda$ & $N_\textrm{s}$ & $D_\textrm{s}/\lambda$ & $N_\textrm{a/s}$ & $D_\textrm{a}/\lambda$ & $f_\textrm{A}$\tnote{b} & $f_\textrm{s}$\tnote{c} \\
\midrule
\texttt{LAMBDA-I} & 2 & $3.9\times 10^6$ & 4 & 17.5 & 256 & 1 &  $1.61\times 10^{-10}$ & 0.836 \\ 
\texttt{SKA-low-core} & 2 & $5.25\times 10^{2}$ & 256 & 17.5 & 256 & 1 & 0.284 & 0.836 \\ 
\texttt{SKA-low} & 2 & $3.5\times 10^4$ & 512 & 17.5 & 256 & 1 &  $1.28\times 10^{-4}$ & 0.836 \\ 
\texttt{CASPA} & 0.25 & $4.6\times 10^4$ & 3 & 8.08 & 65 & 0.5 &  $9.2\times 10^{-8}$ & 0.996 \\ 
\texttt{FarView-core}\tnote{d} & 10 & $6.77\times 10^2$ & 81 & 38.5 & 625 & 1 & 0.262 & 0.422 \\ 
\bottomrule
\end{tabular}
\begin{tablenotes}[hang]
% \item[]Table note
\item[a]Nominal operating wavelength of the array. Results may differ at other wavelengths depending on aperture efficiency, filling factor, etc. 
\item[b]Array filling factor, $\,f_\textrm{A}=N_\textrm{s}\,(D_\textrm{s}/D_\textrm{A})^2$
\item[c]Station filling factor, $\,f_\textrm{s}=N_\textrm{a/s}\,(D_\textrm{a}/D_\textrm{s})^2$
\item[d]Only 6~km core of the proposed array \citep{Polidan+2024} is considered here.
\end{tablenotes}
\end{threeparttable}
\end{table*}

@@

Two broad classes of two-stage fast imaging architectures are considered. In both classes, the first stage is intra-station interferometry of the element electric fields using the options of voltage beamforming (BF), E-field Parallel Imaging Correlator (EPIC), beamforming of correlations (XBF), or correlation followed by FFT (XFFT). The second stage \textbf{involves} inter-station combination of the data products from the first stage. The main difference between the two classes is in whether the second stage uses the first stage products coherently or incoherently. A schematic of these hybrid architectures and their computational cost components are illustrated in Figure~\ref{fig:hybrid-architectures-cost-schematic}.

@@

% \begin{figure*}
% \centering
% \subfloat[][Two-dimensional slices \label{fig:multidim-incoherent-compcost-BF-SKA-low}]{\includegraphics[width=0.4\textwidth]{figures/multidim_beamformer_dft_cost_analysis_SKA1-low.pdf}}
% \subfloat[][One-dimensional slices\label{fig:1D-incoherent-compcost-BF-SKA-low}]{\includegraphics[width=0.58\textwidth]{figures/1D_beamformer_dft_cost_analysis_SKA1-low.pdf}}
% \caption{(Left): Two-dimensional slices of the computational cost for coherent-incoherent imaging using station-level voltage beamforming for \texttt{SKA-low-core}, \texttt{SKA-low}, \texttt{Aus-VLBA-low}, and \texttt{IC-VLBA-low}. Cyan contours  \label{fig:incoherent-compcost-BF-SKA-low}}
% \end{figure*}

@@

Figure~\ref{fig:1D-incoherent-compcost-EPIC-LAMBDA-I} shows a breakdown of the costs per station per voxel of the EPIC architecture, with nominal values of a \texttt{LAMBDA-I} station marked in cyan. The dominant cost is from the two-dimensional FFT (black dashed lines) in Equation~\ref{eqn:intra-station-pol-hol-img-epic}.
% or equivalently, Equation~\ref{eqn:intra-station-pol-hol-img-epic-LA}.
Because the spatial FFT cost depends only on the grid dimensions, it is independent of $N_\textrm{a/s}$ as the grid can hold \textbf{as many} $N_\textrm{a/s}$ elements as physically possible. This spatial FFT scales roughly as $(\gamma_\textrm{s} \, D_\textrm{s}/D_\textrm{a})^2\log_2(\gamma_\textrm{s}\, D_\textrm{s}/D_\textrm{a})^2$, where, $\gamma_\textrm{s}$ is a constant padding factor. Notably, because the spatial FFT produces an image over the entire field of view at every independent pixel instantly, it is not a pixelwise operation, and therefore, its cost per pixel is $\sim \log_2(D_\textrm{s}/D_\textrm{a})$, showing a weak dependence on $D_\textrm{s}$ and $D_\textrm{a}$. The pixelwise squaring (pink dot-dashed lines) and temporal averaging (blue dot-dashed lines) costs in Equation~\ref{eqn:intra-station-opt-pol-img-outprod}
% {eqn:intra-station-opt-polarimetric-direct-img-LA}
are subdominant. These being per pixel operations, the cost per voxel remains constant with $N_\textrm{a/s}$, $D_\textrm{s}$ and $D_\textrm{a}$ similar to the voltage beamforming architecture. The gridding (black dotted lines) with weights in Equation~(\ref{eqn:intra-station-pol-hol-img-epic})
% and (\ref{eqn:intra-station-pol-hol-img-epic-LA})
is a per-element operation and depends only on $N_\textrm{a/s}$, it scales linearly with $N_\textrm{a/s}$ and inverse squared with $D_\textrm{s}$ per voxel. Overall, the gridding cost is insignificant. Again, the spatial averaging of pixels across multiple stations (orange dot-dashed line) is not a station-level operation, but is only shown to emphasise that it operates on a much slower cadence of $t_\textrm{acc}$, scaling inversely with $t_\textrm{acc}$, and adds only negligible cost to the overall cost budget.

@@

\begin{figure}
\includegraphics[width=\linewidth]
% {figures/1D_xcor_dft_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_06.pdf}
\caption{A breakdown of the computational cost density function over the station parameters for the correlator beamforming architecture (XBF) at the station level. Each panel shows the variation with the respective parameter keeping the rest fixed at the characteristic values of the \texttt{LAMBDA-I}, \texttt{SKA-low-core}, and \texttt{SKA-low} stations (cyan lines). The two-dimensional DFT (dashed lines) dominates the cost for the chosen \texttt{LAMBDA-I} parameters at an imaging cadence of $t_\textrm{acc}=1$~ms over correlation (double dot-dashed) and temporal averaging of the correlations (dot-dashed).
% The spatial FFT cost in Equation~(\ref{eqn:intra-station-pol-hol-img-epic-LA}) denoted by black dashed lines dominates the squaring (pink dot-dashes) and temporal averaging (blue dot-dashes) costs in Equation~(\ref{eqn:intra-station-opt-polarimetric-direct-img-LA}). The gridding operation (black dotted lines) in Equation~(\ref{eqn:intra-station-pol-hol-img-epic-LA}) contributes even lesser to the cost budget. 
\label{fig:1D-incoherent-compcost-XBF-LAMBDA-I}}
\end{figure}

@@

\begin{figure*}
\centering
\subfloat[][Covariant cost density of XBF architecture \label{fig:multidim-incoherent-compcost-XBF-LAMBDA-I}]
{\includegraphics[width=0.48\textwidth]
% {figures/multidim_xcor_dft_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_07a.pdf}
}
\subfloat[][Covariant cost of XBF architecture relative to BF architecture \label{fig:multidim-incoherent-relative-compcost-XBF-LAMBDA-I}]
{\includegraphics[width=0.48\textwidth]
% {figures/multidim_xcor_dft_relative_cost_analysis_LAMBDA-I.pdf}
{fig_07b.pdf}
}
\caption{(Left): Two-dimensional slices of the computational cost for coherent-incoherent imaging using a DFT of station-level cross-correlations for \texttt{LAMBDA-I}, \texttt{SKA-low-core}, and \texttt{SKA-low} station parameters (dashed gray lines). Cyan lines denote contours of the colour scale in logarithmic increments. (Right): Same as the left but relative to a voltage beamformer architecture. The computational cost density is dominated by the two-dimensional DFT cost for $t_\textrm{acc}=1$~ms, and is lower than that of voltage beamforming for $N_\textrm{a/s}\lesssim 100$ while getting more expensive with increasing $N_\textrm{a/s}$. \label{fig:incoherent-compcost-XBF-LAMBDA-I}}
\end{figure*}

@@

% \begin{figure*}
% \centering
% \subfloat[][Two-dimensional slices \label{fig:multidim-incoherent-compcost-XBF-SKA-low}]{\includegraphics[width=0.48\textwidth]{figures/multidim_xcor_dft_cost_analysis_SKA1-low.pdf}}
% \subfloat[][Two-dimensional slices (relative)\label{fig:multidim-incoherent-relative-compcost-XBF-SKA-low}]{\includegraphics[width=0.48\textwidth]{figures/multidim_xcor_dft_relative_cost_analysis_SKA1-low.pdf}}
% \caption{(Left): Two-dimensional slices of the computational cost for coherent-incoherent imaging using a DFT of station-level cross-correlations for \texttt{SKA-low-core}, \texttt{SKA-low}, \texttt{Aus-VLBA-low}, and \texttt{IC-VLBA-low}. Cyan contours  \label{fig:incoherent-compcost-XBF-SKA-low}}
% \end{figure*}

% \begin{figure}
% \includegraphics[width=\linewidth]{figures/1D_xcor_dft_cost_analysis_SKA1-low.pdf}
% % \includegraphics[width=\linewidth]{fig_01.pdf}
% \caption{
% \label{fig:1D-incoherent-compcost-XBF-SKA-low}}
% \end{figure}

@@

\begin{figure}
\includegraphics[width=\linewidth]
% {figures/1D_xcor_fft_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_08.pdf}
\caption{A breakdown of the computational cost density function over the station parameters for the correlator FFT architecture (XFFT) at the station level. Each panel shows the variation with the respective parameter keeping the rest fixed at the characteristic values of the \texttt{LAMBDA-I}, \texttt{SKA-low-core}, and \texttt{SKA-low} stations (cyan lines). The correlator / X-engine (double dot- dashed lines) and gridding (dotted lines) dominate the cost for the chosen \texttt{LAMBDA-I} parameters at an imaging cadence of $t_\textrm{acc}=1$~ms over the temporal averaging (dot dahsed lines) and two-dimensional FFT (dashed lines).
\label{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}}
\end{figure}
